RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-09-18 doc/bioinformatics2013/RVD2.tex
updates to hypothesis testing discussion in manuscript

A variant is called if the data indicate that $\mu_j^{\text{case}} > \mu_j^{\text{control}}$ with high confidence,
\begin{equation}
	\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} \geq \tau ) > 1-\alpha,
\end{equation}
where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level. While analytical forms for the posterior distributions are not available, hthe M-w-G algorithm does provide samples from those distributions. So, we sample from the posterior distribution $\mu_j^{\Delta} \triangleq \mu_j^{\text{case}} - \mu_j^{\text{control}}$ by randomly sampling with replacement from the samples of $\mu_j^{\text{case}}$ and$\mu_j^{\text{control}}$ and calculating samples from $\mu_j^{\Delta}$ directly.

@@

We assume that at one position in one experimental replicate, the non-reference read count follows a multinomial distribution of $k = 3$ categories, with probability vector $p = (p_1, p_2, p_3)$ where $p_i \geq 0$ for $i =1, 2, 3$ and $\sum_{i=1}^3 p_i = 1$.

The non-reference reads can be because of either mutation or various errors including sampling error or inter-replicated error. We assume, non-reference read counts caused by these random mechanisms should be uniformly distributed among three non-reference categories, giving $p_1=p_2=p_3=1/3$. In contrast, the distribution of non-reference read counts among three non-reference categories caused by mutation would be biased to one base.

We use a $\chi^2$ goodness-of-fit test to distinguish these two possible scenarios. The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$. Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistics, the log likelihood ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$. The test statistic is
\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
\end{equation}
where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis. Cressie and Read (1984) proved that when $\lambda = 2/3$ this statistic is XXX.

For each position, there are $N$ power-divergence statistics for each experimental replicate. We use Fisher's combined probability test to combine the p-values from each replicates into a single one for each position. 
\begin{equation}
	X^2 = -2 \sum_{i=1}^N \ln(p_i)
\end{equation}

To control for Bonferroni correction is applied to maintaining the familywise error rate (FWER). The significance level is chosen as 0.05.

Finally, if read counts in one position passes the Bayesian Hypothesis Testing and fails the Goodness-of-fit Testing, the position will be called.

For each position we have six power-divergence statistics from 6 experimental replicates. We use Fisher's method to combines the six statistics into a single one for inference. Bonferroni correction is applied to maintaining the familywise error rate (FWER). The significance level is chosen as 0.05.

>>>>

A variant is called if the data indicate that $\mu_j^{\text{case}} > \mu_j^{\text{control}}$ with high confidence,
\begin{equation}
	\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} \geq \tau ) > 1-\alpha,
\end{equation}
where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level. We draw a sample from the posterior distribution $\mu_j^{\Delta} \triangleq \mu_j^{\text{case}} - \mu_j^{\text{control}}$ by randomly sampling with replacement from the samples of $\mu_j^{\text{case}}$ and$\mu_j^{\text{control}}$ and calculating samples from $\mu_j^{\Delta}$ directly.

@@

We assume the counts for the non-reference bases at poisiuton $j$ in replicate $i$ follow a multinomial distribution with parameter  $p = (p_1, p_2, p_3)$ where $p_k \geq 0$ for $k =1, 2, 3$ and $\sum_{k=1}^3 p_k = 1$.

The non-reference reads may be due to a true mutation or due to a random sequencing error. We assume, non-reference read counts caused by a random mechanism results in a uniform distributed among three non-reference categories, giving $p_1=p_2=p_3=1/3$. In contrast, the distribution of non-reference read counts among three non-reference categories caused by mutation would be biased to one base.

We use a $\chi^2$ goodness-of-fit test to distinguish these two possible scenarios. The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$. Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistics, the log likelihood ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$. The test statistic is

\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
\end{equation}

where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis. Cressie and Read (1984) proved that when $\lambda = 2/3$ this statistic is XXX.

For each position, there are $N$ power-divergence statistics for each experimental replicate. We use Fisher's combined probability test to combine the p-values from each replicates into a single one for each position,

\begin{equation}
	X^2 = -2 \sum_{i=1}^N \ln(p_i).
\end{equation}
This gives a test statistic that follows a $\chi^2$ distribution with $2*N$ degrees of freedom when the null hypothesis is true.

While Fisher's combined probability test collapses the p-values across replicates, we still have $J$ tests across positions. We use the Bejamini-Hochberg method to control the family-wise error rate (FWER) across positions.
